---
output: github_document
bibliography: bibliography.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stmoran

<!-- badges: start -->

<!-- badges: end -->

The goal of stmoran is to make the spatio-temporal variants of global and local Moran's proposed by [@gao_measuring_2019] (spatio-temporal LISA) available in R. It borrows code from and works as an extension to Roger Bivand's `spdep` package. I put this together because I needed it: it has not undergone extensive testing and is likely to contain bugs and errors. Please let me know if you encounter any.

## Installation

You can install the development version of stmoran from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("nbtjaden/stmoran")
```

You will also need a working installation of `spdep`:

``` r
install.packages("spdep")
```

## Usage

The functions `moran.test_st` and `localmoran_st` are the spatio-temporal equivalents of `spdep::moran.test` and `spdep::localmoran`. The main difference in using them is that `x` is expected to be a matrix/data.frame of time series of observations at different locations (instead of just a vector of observations at different locations). Load the package with:

```{r example}
library(stmoran)
```

Please refer to `stmoran.html` in the `./vignettes/` directory for an example on how to use `stmoran`

## References
