---
title: "stmoran"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{stmoran}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Setup

## Load Packages

Load `spdep` and `stmoran` packages. Also load `cluster` to get the `votes.repub` dataset and `spData` to get a polygon map of US states.

Load packages:

-   `spdep` - is needed for `stmoran` to work at all
-   `stmoran` - this package
-   `cluster` - provides the `votes.repub` dataset we are going to use as an example
-   `spData` - provides polygons of US states for calculating weights and plotting
-   `dplyr` - is used for data wrangling
-   `ggplot2` - is used for plotting the results
-   `gridExtra` - is used to arrange multiple plots via `grid.arrange()`

```{r setup}
library(spdep)
library(stmoran)
library(cluster)
library(spData)
library(dplyr)
library(ggplot2)
library(gridExtra)
```

## Plot settings

### Colors

Define some colors we might want to use for plotting:

```{r}
myblue1 <- "#c2d7e4"
myblue2 <- "#3f85ad"
myred1 <- "#ff977c"
myred2 <- "#ff4a2c"
mypurple1 <- "#c2a5cf"
mygreen1 <- "#a6dba0"
NAgrey <- "grey80"
```

### ggplot2 helper functions

Pre-define some plotting options to keep the code clean.

```{r}
# global plot options
clearmap <- theme_bw()+
theme(panel.grid = element_line(colour = "transparent"),
axis.title.x=element_blank(),
axis.text.x=element_blank(),
axis.ticks.x=element_blank(),
axis.title.y=element_blank(),
axis.text.y=element_blank(),
axis.ticks.y=element_blank()
)

# plot votes for a certain year
plot_votes <- function(x, title){
  ggplot() +
    geom_sf(data=votes,
            aes(fill = x),
            col="white")+
    scale_fill_viridis_c(name="% votes for republicans ")+
    ggtitle(title)+
    clearmap
}

# plot local Moran's I values
plot_Ii <- function(x, title){
  ggplot() +
    geom_sf(data=votes,
            aes(fill = x),
            col="white")+
    scale_fill_viridis_c(name="Ii value",
                         limits = c(-0.1, 5))+
    ggtitle(title)+
    clearmap
}

# plot significance of local Moran's I
plot_significance <- function(x, title){
  ggplot() +
    geom_sf(data=votes,
            aes(fill = x),
            col="white")+
    scale_fill_manual(values = c(mypurple1, mygreen1),
                      na.value = NAgrey,
                      name = "Ii significant")+
    ggtitle(title)+
    clearmap
}

# plot quadrants for local Moran's I
plot_quadrants <- function(x, title){
  ggplot() +
    geom_sf(data=votes,
            aes(fill = x),
            col="white")+
    scale_fill_manual(values = c("Low-Low"=myblue2, "Low-High"=myblue1, "High-High"=myred2, "High-Low"=myred1),
                      na.value = NAgrey,
                      name = "quadrants")+
    ggtitle(title)+
    clearmap
}

# plot only significant quadrants
plot_quadrants_sigonly <- function(x, y, title){
  ggplot() +
    geom_sf(data=votes,
            aes(fill = x),
            col="white")+
    scale_fill_manual(values = c("Low-Low"=myblue2, "Low-High"=myblue1, "High-High"=myred2, "High-Low"=myred1, "n.s."=NAgrey),
                      na.value = NAgrey,
                      name = "quadrants")+
    ggtitle(title)+
    clearmap
}
```

# Get Data

## US Election Data

Load data about the US presidential elections 1856-1976 per state, expressed as percentage of votes for the republican party:

```{r}
data(votes.repub)
dim(votes.repub)
votes.repub[1:5, 1:5]
```

Extract the column names (years) for later:

```{r}
years <- names(votes.repub)
```

Explore the included US states:

```{r}
row.names(votes.repub)
```

Handling of NA is currently ToDo, for now we will just treat them as undecided and fill the missing data with 50%.

```{r}
votes.repub[is.na(votes.repub)] <- 50
```

## US State Polygons

Explore the polygon data:

```{r}
str(us_states)
```

Basic plot:

```{r}
plot(us_states)
```

Note that the included States are somewhat different (Alaska is missing, for example):

```{r}
us_states$NAME
```

# Data Processing

## Merge Election Data and Polygons

Add state names as a data column, as well as a long-term mean value:

```{r}
votes <- cbind(NAME=row.names(votes.repub),
               votes.repub,
               mean=rowMeans(votes.repub))
```

Merge with the spatial data, doing an inner join to get rid of any states that are not available in both data sets:

```{r}
votes <- inner_join(x = us_states, y = votes, by = "NAME")
str(votes)
```

Plot example map for the 1976 election:

```{r}
plot_votes(votes$"X1976", "Voting for 1976")
```

Plot a map of the 1856-1976 long-term average of election results

```{r}
plot_votes(votes$mean, "1856-1976 long-term average of election results")
```

We will also need a simple data frame for the main calculations. Use `t()` to transpose, i.e. switch rows and columns:

```{r}
votes_simple <- st_drop_geometry(votes[years]) %>% t() %>% data.frame()
```

## Spatial weights

Spatial weights are calculated with standard `spdep` functions (see `spdep` documentation for details).

```{r}
neighbours <- spdep::poly2nb(votes, queen=TRUE)
weightslist <- spdep::nb2listw(neighbours=neighbours, style="W", zero.policy = TRUE)
```

# Global Moran's I

Regular global Moran's I calculated for the long-term mean of the entire period:

```{r}
I_mean <- spdep::moran.test(x = votes$mean, listw = weightslist)
I_mean
```

Spatio-temporal global Moran's I following Gao et al. (2019), based on the time series:

```{r}
I_st <- moran.test_st(x = votes_simple, listw = weightslist)
I_st
```

In both cases, positive I value with low p-values suggest the existence of spatial clusters. The higher I value of `I_st` shows that spatial clustering is stronger when temporal dynamics are being taken into account.

# Local Moran's I (LISA)

Calculate regular local Moran's I based on the long-term means of the entire study period:

```{r}
Ii_mean <- spdep::localmoran(x = votes$mean, listw = weightslist)
```

Calculate spatio-temporal local Moran's I following Gao et al. (2019), based on the time series:

```{r}
Ii_st <- localmoran_st(x = votes_simple, listw = weightslist)
```

Extract the raw Ii values:

```{r}
Ii_plt_mean <- data.frame(Ii_mean)$Ii
Ii_plt_st <- data.frame(Ii_st)$Ii
```

Compare the results:

```{r}
grid.arrange(plot_Ii(Ii_plt_mean, title="1856-1976 mean data"),
             plot_Ii(Ii_plt_st, title="1856-1976 time series data"))
```

From a glimpse at the maps, there seems to be some clustering going on in the South-East and maybe some more in the North-East. Keep in mind that interpretability of Ii values in non-significant areas is questionable. Let's find those areas, using a significance level of p=0.05:

```{r}
siglevel=0.05
sig_mean <- Ii_mean[, ncol(Ii_mean)] <= siglevel
sig_st <- Ii_st[, ncol(Ii_st)] <= siglevel
```

```{r}
grid.arrange(plot_significance(sig_mean, title="1856-1976 mean data"),
            plot_significance(sig_st, title="1856-1976 time series data"))
```

So there seems to be *significant* clustering going on in the South-East and, to a lesser degree in the North-East. That doesn't tell us much about the underlying data, though. The `spdep::localmoran()` function as well as `localmopran_st()` already the information in which quadrant of the *Moran scatterplot* a data point is located. Extract that data:

```{r}
quadr_mean <- attr(Ii_mean, "quadr")
quadr_mean <- ordered(quadr_mean[,"mean"], levels=c("Low-Low", "Low-High", "High-High", "High-Low"))
quadr_st <- attr(Ii_st, "quadr")
quadr_st <- ordered(quadr_st[,"quadr"], levels=c("Low-Low", "Low-High", "High-High", "High-Low"))
```

Plot it:

```{r}
grid.arrange(plot_quadrants(quadr_mean, title="1856-1976 mean data"),
             plot_quadrants(quadr_st, title="1856-1976 time series data"))
```

In the South-East, there is a large Low-Low cluster (low values surrounded by other low values). In this case, both maps are identical, suggesting that time does not play a major role for clustering in this data set. Remember, however, that the maps of statistical significance above showed some differences. It would probably be a good idea to mask out non-significant states to avoid interpreting non-significant results.

Add a new level to the \`quadr\` factors for "not significant:

```{r}
quadr_mean_sigonly <- quadr_mean
levels(quadr_mean_sigonly) <- c(levels(quadr_mean_sigonly), "n.s.")
quadr_mean_sigonly[!sig_mean] <- "n.s."

quadr_st_sigonly <- quadr_st
levels(quadr_st_sigonly) <- c(levels(quadr_st_sigonly), "n.s.")
quadr_st_sigonly[!sig_st] <- "n.s."
```

Plot the maps again, this time only showing significant clusters:

```{r}
grid.arrange(plot_quadrants_sigonly(x=quadr_mean_sigonly, title="1856-1976 mean data"),
             plot_quadrants_sigonly(x=quadr_st_sigonly, title="1856-1976 time series data"))
```

Now it is much easier to see that there is a large cluster of disapproval for the republican party in the South-East and a smaller cluster of high approval in the North-East. However, the spatio-temporal analyses suggests that these clusters might be smaller than assumed when only using long-term average data. Especially in the North, they may not be persistent in time.
